\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{enumitem}

\setlist[itemize]{leftmargin=1.5em}
\setlist[enumerate]{leftmargin=1.8em}

\newcommand{\placeholder}[1]{\textit{[#1]}}

\title{CS336 Assignment 1 (Basics)\\Writeup}
\author{Lenard Strahringer}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\newpage

\section{Problem (unicode1): Understanding Unicode}
\subsection*{(a)}
Let \(B=\text{batch\_size}\), \(V=\text{vocab\_size}\), \(T=\text{context\_length}\), \(L=\text{num\_layers}\), \(d=d_{\text{model}}\), \(h=\text{num\_heads}\), \(d_{\text{ff}}=\frac{8}{3}d\). Float32 uses 4 bytes/element.

\[
P = 2Vd + L\!\left(4d^2 + 3dd_{\text{ff}} + 2d\right) + d
:= 2Vd + L(12d^2 + 2d) + d .
\]

\[
M_{\text{params}} = 4P \text{ bytes}.
\]

\[
\begin{aligned}
A_{\text{block}} &= \underbrace{2BTd}_{\text{2 RMSNorms}}
+ \underbrace{(3BTd + BhT^2 + BhT^2 + BTd + BTd)}_{\text{QKV, }QK^\top,\text{ softmax, weighted sum, }W_O}
+ \underbrace{(BTd_{\text{ff}} + BTd_{\text{ff}} + BTd_{\text{ff}} + BTd_{\text{ff}} + BTd)}_{\text{W1, W3, SiLU, product, W2}} \\
&= 8BTd + 4BTd_{\text{ff}} + 2BhT^2
:= \frac{56}{3}BTd + 2BhT^2 .
\end{aligned}
\]

\[
A_{\text{extra}} = BTd + 2BTV.
\]

\[
A = L A_{\text{block}} + A_{\text{extra}}
:= L\!\left(\frac{56}{3}BTd + 2BhT^2\right) + BTd + 2BTV .
\]

\[
M_{\text{act}} = 4A \text{ bytes}.
\]

\[
M_{\text{grad}} = 4P \text{ bytes}.
\]

\[
M_{\text{opt}} = 8P \text{ bytes}.
\]

\[
M_{\text{peak}} = M_{\text{params}} + M_{\text{act}} + M_{\text{grad}} + M_{\text{opt}}
:= 16P + 4A \text{ bytes}.
\]
\[
\boxed{
M_{\text{peak}}
= 32\,\text{vocab\_size}\,d_{\text{model}}
+ 192\,\text{num\_layers}\,d_{\text{model}}^2
+ 32\,\text{num\_layers}\,d_{\text{model}}
+ 16\,d_{\text{model}}
+ \frac{224}{3}\,\text{num\_layers}\,\text{batch\_size}\,\text{context\_length}\,d_{\text{model}}
+ 8\,\text{num\_layers}\,\text{batch\_size}\,\text{num\_heads}\,\text{context\_length}^2
+ 4\,\text{batch\_size}\,\text{context\_length}\,d_{\text{model}}
+ 8\,\text{batch\_size}\,\text{context\_length}\,\text{vocab\_size}
\;\text{bytes}.}
\]

\subsection*{(b)}
Its string representation shows an escape sequence like \texttt{'\textbackslash x00'}, while printing it produces a non-printing control character (typically visually empty output).

\subsection*{(c)}
When this character appears in text it is embedded as an invisible null byte that still occupies a position in the string, and some tools/languages may treat it as a terminator and behave unexpectedly.

\section{Problem (unicode2): Unicode Encodings}
\subsection*{(a)}
UTF-8 is usually preferred because it is much more space-efficient for ASCII-heavy text (common on the web), is byte-oriented and backward-compatible with ASCII, and avoids UTF-16 endianness/BOM complications while still representing all Unicode characters.

\subsection*{(b)}
An example is \texttt{b'\textbackslash xc3\textbackslash xa9'} (which is ``é'' in UTF-8): the function is incorrect because it decodes each byte independently, but UTF-8 is variable-length and many characters require combining multiple bytes before decoding.

\subsection*{(c)}
An example invalid two-byte sequence is \texttt{b'\textbackslash xc0\textbackslash xaf'}, which is an overlong UTF-8 encoding and therefore does not decode to any valid Unicode character sequence.

\section{Problem (train\_bpe\_tinystories): BPE Training on TinyStories}
\subsection*{(a)}
Using \texttt{input\_path=data/TinyStoriesV2-GPT4-train.txt}, \texttt{max\_vocab\_size=10000}, and \texttt{special\_tokens=['<|endoftext|>']}, training took 1017.59 seconds and produced a 10,000-token vocabulary whose longest token was 15 bytes: \texttt{' accomplishment'}. This longest token is reasonable because frequent whitespace-prefixed word fragments are exactly the kind of units BPE tends to merge.

\subsection*{(b)}
In my profile, most runtime is in the BPE merge loop: \texttt{update\_word\_pairs} (32.38\%) and \texttt{merge\_tokens\_words\_element} (32.16\%) together account for about two-thirds of sampled execution time. Smaller portions are loop overhead and preprocessing/counting, and because subprocess profiling was not enabled, worker-process pretokenization time may be underrepresented.

\section{Problem (train\_bpe\_expts\_owt): BPE Training on OpenWebText}
\subsection*{(a)}
Using \texttt{input\_path=data/owt\_valid.txt}, \texttt{max\_vocab\_size=32000}, and \texttt{special\_tokens=['<|endoftext|>']}, training took 1943.20 seconds and produced a 32,000-token vocabulary whose longest token was 64 bytes: \texttt{'----------------------------------------------------------------'}. This token is plausible because long repeated punctuation spans occur in web text and are highly mergeable under BPE.

\subsection*{(b)}
\placeholder{One-to-two sentence compare/contrast with TinyStories tokenizer.}

\section{Problem (tokenizer\_experiments): Experiments with Tokenizers}
\subsection*{(a)}
On 10 sampled TinyStories documents, the TinyStories tokenizer achieved 4.150 bytes/token (7412 bytes over 1786 tokens). On 10 sampled OpenWebText documents, the OpenWebText tokenizer achieved 4.680 bytes/token (31487 bytes over 6728 tokens), indicating better compression on its in-domain sample.

\subsection*{(b)}
Tokenizing the OpenWebText sample with the TinyStories tokenizer produced 3.160 bytes/token (31487 bytes over 9963 tokens), versus 4.680 bytes/token with the OpenWebText tokenizer on the same sample. This means the TinyStories tokenizer produced many more tokens (worse compression), with a cross-minus-native difference of $-1.520$ bytes/token.

\subsection*{(c)}
Measured throughput was about 4.12 MB/s for the TinyStories tokenizer and 1.92 MB/s for the OpenWebText tokenizer (benchmarked on about 5.0M bytes each). At those rates, tokenizing the 825GB Pile is estimated to take about 55.67 hours (2.32 days) with the TinyStories tokenizer or 119.08 hours (4.96 days) with the OpenWebText tokenizer.

\subsection*{(d)}
\texttt{uint16} is appropriate because both vocabularies are below 65,536 entries (10K and 32K), so every token ID fits exactly in 16 bits. Using \texttt{uint16} therefore halves storage relative to \texttt{int32} while preserving all IDs.

\section{Problem (transformer\_accounting): Transformer LM Resource Accounting}
\subsection*{(a)}
Trainable parameters (no linear biases; \texttt{RMSNorm} has only a scale vector):
\[
\text{total}
= (50{,}257\cdot1{,}600)
+ 48\left(4\cdot1{,}600^2 + 3\cdot1{,}600\cdot4{,}288 + 2\cdot1{,}600\right)
+ 1{,}600
+ (1{,}600\cdot50{,}257)
\]
\[
= 80{,}411{,}200 + 1{,}479{,}628{,}800 + 1{,}600 + 80{,}411{,}200
= 1{,}640{,}452{,}800
\]

If each is stored as a single-precision floating point, the model needs 32 bits x 1{,}640{,}452{,}800 = 52,494,489,600 \text{ bits} \approx 6.56 \text{ GB}. 6.56 \text{ GB} are required just to load the model.

\subsection*{(b)}
For one \texttt{TransformerBlock}, let \(B=\) batch size, \(T=\) sequence length, \(d=1600\), \(h=25\), \(d_k=d/h=64\), \(d_{\text{ff}}=4288\). The 9 matrix multiplications are:
\begin{itemize}
    \item Q projection: \([B,T,d]\times[d,d]\to[B,T,d]\)
    \item K projection: \([B,T,d]\times[d,d]\to[B,T,d]\)
    \item V projection: \([B,T,d]\times[d,d]\to[B,T,d]\)
    \item Attention scores \(QK^\top\): \([B,h,T,d_k]\times[B,h,d_k,T]\to[B,h,T,T]\)
    \item Attention-value product: \([B,h,T,T]\times[B,h,T,d_k]\to[B,h,T,d_k]\)
    \item Output projection \(W_O\): \([B,T,d]\times[d,d]\to[B,T,d]\)
    \item FFN branch \(W_1\): \([B,T,d]\times[d,d_{\text{ff}}]\to[B,T,d_{\text{ff}}]\)
    \item FFN branch \(W_3\): \([B,T,d]\times[d,d_{\text{ff}}]\to[B,T,d_{\text{ff}}]\)
    \item FFN output \(W_2\): \([B,T,d_{\text{ff}}]\times[d_{\text{ff}},d]\to[B,T,d]\)
\end{itemize}

For \texttt{TransformerLM} outside the \texttt{TransformerBlock}s:
\begin{itemize}
    \item Final output projection: \([B,T,d]\times[d,\text{vocab\_size}]\to[B,T,\text{vocab\_size}]\)
\end{itemize}

One forward pass needs \(9\times48+1=433\) matrix multiplications.

General rule for matrix multiplication FLOPs:
\[
([m,n]\times[n,p]) \Rightarrow 2mnp \text{ FLOPs}
\]

Using \(B=1\), \(T=1024\), \(d=1600\), \(h=25\), \(d_k=64\), \(d_{\text{ff}}=4288\), \(\text{vocab\_size}=50{,}257\):
\begin{itemize}
    \item Q, K, V, \(W_O\): each \(2BTdd=5{,}242{,}880{,}000\)
    \item \(QK^\top\), attention-value: each \(2BhTTd_k=3{,}355{,}443{,}200\)
    \item \(W_1, W_3, W_2\): each \(2BTdd_{\text{ff}}=14{,}050{,}918{,}400\)
\end{itemize}

Per-block FLOPs:
\[
4(5{,}242{,}880{,}000) + 2(3{,}355{,}443{,}200) + 3(14{,}050{,}918{,}400)
= 69{,}835{,}161{,}600
\]

All 48 blocks:
\[
48\times69{,}835{,}161{,}600
= 3{,}352{,}087{,}756{,}800
\]

Final output projection FLOPs:
\[
2BTd\cdot\text{vocab\_size}
= 2\cdot1\cdot1024\cdot1600\cdot50{,}257
= 164{,}682{,}137{,}600
\]

Total FLOPs in one forward pass:
\[
3{,}352{,}087{,}756{,}800 + 164{,}682{,}137{,}600
= \boxed{3{,}516{,}769{,}894{,}400}
\]

\subsection*{(c)}
The PositionwiseFeedForward requires a lot of FLOPs (14,050,918,400 x 3 = 42,152,755,200) which is more than 4 times more than the CausalMultiHeadSelfAttention (3,355,443,200 * 2 + 5,242,880,000 = 11,953,766,400).

\subsection*{(d)}
Using the same accounting as above the FLOP proportions are:

\begin{center}
\begin{tabular}{lrrrrrrr}
\toprule
Model & QKV & Attn scores & Attn value & Attn out & FFN W1/W3 & FFN W2 & Final proj \\
\midrule
GPT-2 small  & 14.91\% & 6.63\% & 6.63\% & 4.97\% & 26.51\% & 13.25\% & 27.10\% \\
GPT-2 medium & 18.62\% & 6.21\% & 6.21\% & 6.21\% & 33.37\% & 16.68\% & 12.70\% \\
GPT-2 large  & 20.49\% & 5.46\% & 5.46\% & 6.83\% & 36.20\% & 18.10\% & 7.45\% \\
GPT-2 XL     & 21.47\% & 4.58\% & 4.58\% & 7.16\% & 38.36\% & 19.18\% & 4.68\% \\
\bottomrule
\end{tabular}
\end{center}

As model size increases, FFN and projection-heavy terms take a larger share of total FLOPs, while the final vocab projection takes a much smaller share. The attention score/value terms decrease proportionally here because \(T\) is fixed and \(d\)-/depth-driven terms grow faster.

\subsection*{(e)}
For GPT-2 XL, increasing context length from \(1024\) to \(16{,}384\) raises one-forward-pass matmul FLOPs from \(3{,}516{,}769{,}894{,}400\) to \(133{,}577{,}729{,}638{,}400\), i.e. \(\approx 37.98\times\) higher. The relative FLOP share shifts strongly toward the \(T^2\) attention terms: attention scores + attention-value rise from \(9.16\%\) to \(61.73\%\), while FFN (\(W_1/W_3/W_2\)) falls from \(57.51\%\) to \(24.24\%\) and final projection falls from \(4.68\%\) to \(1.97\%\).

\section{Problem (learning\_rate\_tuning): Tuning the Learning Rate (SGD toy example)}
With 1e1 it decreases over time. With 1e2 it decreases faster. With 1e3 it diverges.

\section{Problem (adamw\_accounting): Resource Accounting for Training with AdamW}
\subsection*{(a)}
Let \(B=\text{batch\_size}\), \(V=\text{vocab\_size}\), \(T=\text{context\_length}\), \(L=\text{num\_layers}\), \(d=d_{\text{model}}\), \(h=\text{num\_heads}\), \(d_{\text{ff}}=\frac{8}{3}d\). Float32 uses 4 bytes/element.

\[
P = 2Vd + L\!\left(4d^2 + 3dd_{\text{ff}} + 2d\right) + d
= 2Vd + L(12d^2 + 2d) + d .
\]

\[
M_{\text{params}} = 4P \text{ bytes}.
\]


\[
\begin{aligned}
A_{\text{block}} &= \underbrace{2BTd}_{\text{2 RMSNorms}}
+ \underbrace{(3BTd + BhT^2 + BhT^2 + BTd + BTd)}_{\text{QKV, }QK^\top,\text{ softmax, weighted sum, }W_O}
+ \underbrace{(BTd_{\text{ff}} + BTd_{\text{ff}} + BTd_{\text{ff}} + BTd_{\text{ff}} + BTd)}_{\text{W1, W3, SiLU, product, W2}} \\
&= 8BTd + 4BTd_{\text{ff}} + 2BhT^2
= \frac{56}{3}BTd + 2BhT^2 .
\end{aligned}
\]

\[
A_{\text{extra}} = BTd + 2BTV.
\]

\[
A = L A_{\text{block}} + A_{\text{extra}}
= L\!\left(\frac{56}{3}BTd + 2BhT^2\right) + BTd + 2BTV .
\]

\[
M_{\text{act}} = 4A \text{ bytes}.
\]


\[
M_{\text{grad}} = 4P \text{ bytes}.
\]


\[
M_{\text{opt}} = 8P \text{ bytes}.
\]


\[
M_{\text{peak}} = M_{\text{params}} + M_{\text{act}} + M_{\text{grad}} + M_{\text{opt}}
= 16P + 4A \text{ bytes},
\]
\[
\boxed{
M_{\text{peak}}
= 32\,\text{vocab\_size}\,d_{\text{model}}
+ 192\,\text{num\_layers}\,d_{\text{model}}^2
+ 32\,\text{num\_layers}\,d_{\text{model}}
+ 16\,d_{\text{model}}
+ \frac{224}{3}\,\text{num\_layers}\,\text{batch\_size}\,\text{context\_length}\,d_{\text{model}}
+ 8\,\text{num\_layers}\,\text{batch\_size}\,\text{num\_heads}\,\text{context\_length}^2
+ 4\,\text{batch\_size}\,\text{context\_length}\,d_{\text{model}}
+ 8\,\text{batch\_size}\,\text{context\_length}\,\text{vocab\_size}
\;\text{bytes}.}
\]

\subsection*{(b)}
Using \(V=50{,}257\), \(T=1024\), \(L=48\), \(d=1600\), \(h=25\), \(d_{\text{ff}}=4288\):
\[
M_{\text{peak}}(B)=aB+b
=16{,}373{,}391{,}360\cdot B + 26{,}247{,}244{,}800
\quad\text{bytes}.
\]
Under an 80GB budget (\(80{,}000{,}000{,}000\) bytes), the largest integer batch size is
\[
B_{\max}=\left\lfloor\frac{80{,}000{,}000{,}000-26{,}247{,}244{,}800}{16{,}373{,}391{,}360}\right\rfloor=3.
\]
Check: \(M_{\text{peak}}(3)=75{,}367{,}418{,}880<80{,}000{,}000{,}000\), while \(M_{\text{peak}}(4)=91{,}740{,}810{,}240>80{,}000{,}000{,}000\).

\subsection*{(c)}
\begin{align*}
    \text{1. Weight decay:} \quad & p \gets p - (\eta \lambda) p && \text{(2 FLOPs)} \\
    \text{2. First moment ($m$):} \quad & m \gets \beta_1 m + (1 - \beta_1) g && \text{(3 FLOPs)} \\
    \text{3. Second moment ($v$):} \quad & v \gets \beta_2 v + (1 - \beta_2) g^2 && \text{(4 FLOPs)} \\
    \text{4. Parameter update:} \quad & p \gets p - \alpha_t \frac{m}{\sqrt{v} + \epsilon} && \text{(5 FLOPs)} \\[1ex]
    \textbf{Total:} \quad & \textbf{14 FLOPs per parameter} \implies \mathbf{14P \text{ FLOPs}} &&
\end{align*}

\subsection*{(d)}
\[
202.086 \text{ days} \times 24 \text{ hours/day} \approx \mathbf{4,850 \text{ hours}}.
\]
Justification: from part (transformer\_accounting), one forward pass for one sequence costs \(3{,}516{,}769{,}894{,}400\) FLOPs. Multiplying by \(\text{batch\_size}\) gives forward FLOPs per step, and using \(\approx 3\times\) forward for forward+backward gives training FLOPs per step. At 50\% MFU on one H100, effective throughput is \(495\times 10^{12}\times 0.5\) FLOPs/s, so training time is total FLOPs divided by this rate.

\section{Problem (training\_together): Put it Together}
The script for Question 5.3 (training loop) is in \texttt{cs336\_basics/train\_lm.py}.

\section{Problem (decoding): Decoding}
The script for Question 6.1 (decoding/generation) is in \texttt{cs336\_basics/train\_lm.py}.

\section{Problem (experiment\_log): Experiment Logging}
For Question 7.1, experiment logging is implemented in \texttt{cs336\_basics/run\_logged\_experiment.py} and \texttt{cs336\_basics/plot\_experiment\_log.py}, with core train-loop logging in \texttt{cs336\_basics/train\_lm.py}.

\section{Problem (learning\_rate): Tune the Learning Rate on TinyStories}
\subsection*{(a)}
\placeholder{Include learning curves for multiple learning rates and explain search strategy.}

\subsection*{(Validation target)}
\placeholder{Report model achieving TinyStories validation loss target (or adjusted low-resource target, if applicable).}

\subsection*{(b)}
\placeholder{Include curves showing increasing learning rates with at least one divergent run + analysis.}

\section{Problem (batch\_size\_experiment): Batch Size Variations}
\placeholder{Include learning curves for multiple batch sizes (including 64 and 128) with retuned learning rates if needed.}

\placeholder{A few sentences discussing findings on batch size effects.}

\section{Problem (generate): Generate Text}
\placeholder{Text dump of at least 256 tokens (or until first \texttt{<|endoftext|>}).}

\placeholder{Brief comment on fluency and at least two factors affecting output quality.}

\section{Problem (layer\_norm\_ablation): Remove RMSNorm and Train}
\placeholder{Learning curve(s) for no-RMSNorm model, including best lower-learning-rate run.}

\placeholder{A few sentences on RMSNorm impact.}

\section{Problem (pre\_norm\_ablation): Implement Post-Norm and Train}
\placeholder{Learning curve comparing post-norm and pre-norm Transformers.}

\section{Problem (no\_pos\_emb): Implement NoPE}
\placeholder{Learning curve comparing RoPE and NoPE.}

\section{Problem (swiglu\_ablation): SwiGLU vs.\ SiLU}
\placeholder{Learning curve comparing SwiGLU and SiLU feed-forward networks (parameter-matched).}

\placeholder{A few sentences discussing findings.}

\section{Problem (main\_experiment): Experiment on OpenWebText}
\placeholder{Learning curve on OpenWebText + discussion of loss differences vs TinyStories.}

\placeholder{Generated text in TinyStories-style format + discussion of fluency and quality gap.}

\section{Problem (leaderboard): Leaderboard Submission}
\placeholder{Final validation loss, learning curve with wall-clock x-axis under 45 minutes, and description of modifications/hyperparameters.}

\end{document}
